Ultrafast

### 2023-07-29 manuscript/manuscript.qmd
Add conda env to paper generation

We used four separate datasets for this study. The first is a set of azaphenalenes previously studied by some of us, comprising 257 substituted compounds, for which excitation energies were computed with (2nd-order) algebraic-diagrammatic construction, {{< var reference_level >}}/[Robert: Fill out], and geometries were optimized with B3LYP/[Robert: Fill out][@polliceOrganicMoleculesInverted2021] The second is a set substituted 138 aza-azulenes with {{< var reference_level >}}/[Robert: Fill out] excitation energies and optimized with the B97-3c composite method.[Robert: Add ref to dataset if it exists.] The third is a set of 16 rationally designed scaffolds, which including substitutions amounts to {{< var global_n_total >}} unique compounds, optimized at the B97-3c level and with excitation energies at the {{< var reference_level >}}/[Robert: Fill out] level.[Robert: Add ref to dataset if it exists.] The fourth is a set 315 non-alternant hydrocarbons including substitutions, divided into three subsets of size 76, 187 and 52, optimized at the ωB97X-D/def2-TZVP level and with excitation energies at the CC2/aug-cc-pVTZ level.

@@

We first demonstrate that our method is capable of capturing the inverted $\DeltaE{ST}$ for some model compounds. As shown by Toyota and co-workers, pentalene at the ideal  $D_\mathrm{2h}$ geometry has an inverted gap, while relaxation to the minimum with $C_\mathrm{2h}$ symmetry leads to a normal gap.[@kosekiViolationHundMultiplicity1985] In our calculations, pentalene with equal bond lengths of 1.4 Å displays a small HOMO-LUMO overlap of {{< var pentalene_ideal_overlap >}}, leading to a small exchange interaction of only $2K=$ {{< var pentalene_ideal_2_exchange >}} kcal/mol (@fig-pentalene-ideal-orbitals). The dynamic spin polarization correction $\DeltaE[DSP]{SCF}$ of {{< var pentalene_ideal_dsp_scf >}} kcal/mol leads to a net predicted $\DeltaEST[DSP]{SCF}$ of {{< var pentalene_ideal_t1_s1_dsp_scf >}} kcal/mol. Adding additional correlation with CIS leads to a $\DeltaEST[DSP]{CIS}$ of {{< var pentalene_ideal_t1_s1_dsp_cis >}} kcal/mol. The reference value calculated with {{< var reference_level >}} is {{< var azaphenalene_t1_s1_ref >}} kcal/mol. The simple perturbation theory model allows us to gain additional insight into the contributions to $\DeltaE[DSP]{SCF}$ as they correspond to single excitations between doubly occupied orbitals below the HOMO and unoccupied orbitals above the LUMO (@fig-pentalene-ideal-excitations). Three excitations, HOMO-3 → LUMO+2, HOMO-2 → LUMO+1 and HOMO-1 → LUMO+3 are identified as the main contributors to the spin polarization effect, and could potentially be tuned by substituent effects.

@@

Azaphenalene is the prototypical molecule that started the renewed investigations into INVEST molecules, and has been the subject of numerous studies with high-level quantum chemical methods. Can we capture the inverted gap, as first measured by Leupin and Wirz?[@leupinLowlyingElectronicallyExcited1980] The calculations reveal that the HOMO and LUMO are well separated with spatial overlap of {{< var azaphenalene_overlap >}} and a correspondingly low exchange interaction of {{< var azaphenalene_2_exchange >}} kcal/mol ([@fig-azaphenalene-orbitals]). The larger azaphenalene has 30 excitations that could possibly contribute to the large $\DeltaE[DSP]{SCF}$ of {{< var azaphenalene_dsp_scf >}} kcal/mol, but the most import by far are those from the doubly degenerate HOMO-1 to the doubly degenerate LUMO+1 ([@fig-azaphenalene-excitations]). In total, $\DeltaEST[DSP]{SCF}=$ {{< var azaphenalene_t1_s1_dsp_scf >}} kcal/mol, which is slightly smaller at the CIS level with $\DeltaEST[DSP]{CIS}=$ {{< var azaphenalene_t1_s1_dsp_cis >}} kcal/mol.

@@

While we see success for the azaphenalenes, the situation for the azulenes is unfortunately worse. The PPP level S~1~ and T~1~ excitation energies are only moderately correlated with those from {{< var reference_level >}} with $R^2$ of {{< var local_azulenes_r2_singlet >}} and {{< var local_azulenes_r2_triplet >}}, respectively ([@fig-local-azulenes]a,b). For the oscillator strengths, the situation is rather catastrophic with $R^2$ of {{< var local_azulenes_r2_f >}} implying no correlation whatsoever ([@fig-local-azulenes]c). The situation for the $\DeltaE{ST}$ also worse than for the azaphenalenes, with $R^2=$ {{< var local_azulenes_r2_cis >}} with CIS + DSP, a marked improvement on {{< var local_azulenes_r2_scf >}} with SCF + DSP ([@fig-local-azulenes]e,f). Unfortunately, the negative gaps are not recovered, leading to recalls of {{< var local_azulenes_recall_cis >}} as none of five inverted molecules could be identified ([@tbl-local-azulenes]). It could be speculated that the worse performance for the azulenes comes from the fact that they are non-alternants, for which the approximations are expected to be less applicable. Considering the contrasting performance for the azaphenalenes and azulenes, it seems clear that the PPP screening protocol will inevitably miss some hits belonging to particular compound classes. To investigate the global performance of the model we turned to a more diverse set with 16 different compound classes.

@@

Here we used a dataset previously generated by some of us using rational design rules, comprising 16 different molecular scaffolds, the details of which will be presented elsewhere. Out of {{< var global_n_total >}} compounds, only {{< var global_n_failed >}} failed, which corresponds to a success rate of {{< var global_percent_success >}}. The total runtime using four cores on the M2 processor was {{< var time_cpu_hours >}} CPU hours. The S~1~ and T~1~ excitation energies are fairly reproduced with ([@fig-global-all]a,b) $R^2$ of {{< var global_all_r2_singlet >}} and {{< var global_all_r2_triplet >}} vs {{< var reference_level >}}, respectively. As seen in [@fig-global-all-gaps]a, the $\DeltaE{ST}$ at the SCF + DSP level shows a moderate $R^2$ of {{< var global_all_r2_scf >}}, which increases significantly to {{< var global_all_r2_cis >}} at the CIS + DSP level ([@fig-global-all-gaps]b). However, due to a systematic overestimation of $\DeltaE{ST}$, the F1 score is still low at {{< var global_all_f1_cis >}} and a poor recall of {{< var global_all_recall_cis >}} ([@tbl-global-all]). We therefore added a linear correction $\DeltaEST[DSP]{CIS,LC}=$ {{< var global_lc_coef >}} $\times\DeltaEST[DSP]{CIS}$ {{< var global_lc_intercept >}}, which increases the F1 score to {{< var global_all_f1_cis_corr >}} and the recall to {{< var global_all_recall_cis_corr >}} ([@fig-global-all-gaps]c). The specificity simultaneously decreases from {{< var global_all_specificity_cis >}} to {{< var global_all_specificity_cis_corr >}}, but overall the linear correction would be preferable for virtual screening where there is a stronger focus on finding rare hits. Applying the linear correction to the azaphenalenes ([@fig-local-corr]a) and ([@fig-local-corr]b) leads to an improvement in the recall of the former from {{< var local_azaphenalenes_recall_cis >}} to {{< var local_azaphenalenes_recall_cis_corr >}} while still none of the inverted azulenes are recovered. The oscillator strengths are unfortunately poorly correlated for the rational design set, with $R^2=$ {{< var global_all_r2_oscillator >}}, although the Spearman rank correlation is more encouraging at {{< var global_all_rho_oscillator >}} ([@fig-global-all]c). Taken together, the results reinforce the conclusions from the case study of the azaphenalenes, that the PPP protocol is suitable for screening $\DeltaE{ST}$ while it struggles for oscillator strengths.

A more detailed breakdown of the correlations into the different scaffolds shows that the highest $R^2$ of {{< var global_scaffolds_r2_max >}} is obtained for {{< var global_scaffolds_r2_max_name >}}, while the lowest is obtained for {{< var global_scaffolds_r2_min_name >}} with {{< var global_scaffolds_r2_min >}}. For recall, the corresponding range is {{< var global_scaffolds_recall_max >}} for {{< var global_scaffolds_recall_max_name >}} to {{< var global_scaffolds_recall_min >}} for {{< var global_scaffolds_recall_min_name >}}. The wide range in $R^2$ and recall further reinforces that the method struggles with some particular compounds classes despite its favorable global performance. Despite these problems, the method is able to capture inverted molecules in {{< var global_scaffolds_n_captured >}} out of the 15 scaffolds where they occur. Even for the azulenes included in this screening set at least some candidates are found, in contrast to the findings above for the azaazulenes. The per-scaffold oscillator strengths are arguably sufficiently good to allow for local screening in some cases ([@fig-global-scaffolds-oscillator]). The per-scaffold S~1~ and T~1~ excitation energies can be found in ([@fig-global-scaffolds-singlet]) and ([@fig-global-scaffolds-triplet]), respectively. Notable failures are for {{< var I >}} with $R^2$ of {{< var global_scaffolds_singlet_r2_I >}} and {{< var global_scaffolds_triplet_r2_I >}}, respectively, and {{< var XII >}} and {{< var XVI >}}, where some negative T~1~ excitation energies indicate restricted-unrestricted instabilities.

@@

>>>>

We used four separate datasets for this study. The first is a set of azaphenalenes previously studied by some of us, comprising 257 substituted compounds, for which excitation energies were computed with (2nd-order) algebraic-diagrammatic construction, {{< var reference_level >}}/[Robert: Fill out], and geometries were optimized with B3LYP/[Robert: Fill out][@polliceOrganicMoleculesInverted2021] The second is a set substituted 138 azaazulenes with {{< var reference_level >}}/[Robert: Fill out] excitation energies and optimized with the B97-3c composite method.[Robert: Add ref to dataset if it exists.] The third is a set of 16 rationally designed scaffolds, which including substitutions amounts to {{< var global_n_total >}} unique compounds, optimized at the B97-3c level and with excitation energies at the {{< var reference_level >}}/[Robert: Fill out] level.[Robert: Add ref to dataset if it exists.] The fourth is a set 315 non-alternant hydrocarbons including substitutions, divided into three subsets of size 76, 187 and 52, optimized at the ωB97X-D/def2-TZVP level and with excitation energies at the CC2/aug-cc-pVTZ level.

@@

We first demonstrate that our method is capable of capturing the inverted $\DeltaE{ST}$ for some model compounds. As shown by Toyota and co-workers, pentalene at the ideal  $D_\mathrm{2h}$ geometry has an inverted gap, while relaxation to the minimum with $C_\mathrm{2h}$ symmetry leads to a normal gap.[@kosekiViolationHundMultiplicity1985] In our calculations, pentalene with equal bond lengths of 1.4 Å displays a small HOMO-LUMO overlap of {{< var pentalene_ideal_overlap >}}, leading to a small exchange interaction of only $2K=$ {{< var pentalene_ideal_2_exchange >}} kcal/mol (@fig-pentalene-ideal-orbitals). The dynamic spin polarization correction $\DeltaE[DSP]{SCF}$ of {{< var pentalene_ideal_dsp_scf >}} kcal/mol leads to a net predicted $\DeltaEST[DSP]{SCF}$ of {{< var pentalene_ideal_t1_s1_dsp_scf >}} kcal/mol. Adding additional correlation with CIS leads to a $\DeltaEST[DSP]{CIS}$ of {{< var pentalene_ideal_t1_s1_dsp_cis >}} kcal/mol. The reference value calculated with {{< var reference_level >}} is {{< var pentalene_ideal_t1_s1_ref >}} kcal/mol. The simple perturbation theory model allows us to gain additional insight into the contributions to $\DeltaE[DSP]{SCF}$ as they correspond to single excitations between doubly occupied orbitals below the HOMO and unoccupied orbitals above the LUMO (@fig-pentalene-ideal-excitations). Three excitations, HOMO-3 → LUMO+2, HOMO-2 → LUMO+1 and HOMO-1 → LUMO+3 are identified as the main contributors to the spin polarization effect, and could potentially be tuned by substituent effects.

@@

Azaphenalene is the prototypical molecule that started the renewed investigations into INVEST molecules,[@desilvaInvertedSingletTriplet2019;@sobolewskiAreHeptazineBasedOrganic2021] and has been the subject of numerous studies with high-level quantum chemical methods. Can we capture the inverted gap, as first measured by Leupin and Wirz?[@leupinLowlyingElectronicallyExcited1980] The calculations reveal that the HOMO and LUMO are well separated with spatial overlap of {{< var azaphenalene_overlap >}} and a correspondingly low exchange interaction of {{< var azaphenalene_2_exchange >}} kcal/mol ([@fig-azaphenalene-orbitals]). The larger azaphenalene has 30 excitations that could possibly contribute to the large $\DeltaE[DSP]{SCF}$ of {{< var azaphenalene_dsp_scf >}} kcal/mol, but the most import by far are those from the doubly degenerate HOMO-1 to the doubly degenerate LUMO+1 ([@fig-azaphenalene-excitations]). In total, $\DeltaEST[DSP]{SCF}=$ {{< var azaphenalene_t1_s1_dsp_scf >}} kcal/mol, which is slightly smaller at the CIS level with $\DeltaEST[DSP]{CIS}=$ {{< var azaphenalene_t1_s1_dsp_cis >}} kcal/mol.

@@

While we see success for the azaphenalenes, the situation for the azaazulenes is unfortunately worse. The PPP level S~1~ and T~1~ excitation energies are only moderately correlated with those from {{< var reference_level >}} with $R^2$ of {{< var local_azulenes_r2_singlet >}} and {{< var local_azulenes_r2_triplet >}}, respectively ([@fig-local-azulenes]a,b). For the oscillator strengths, the situation is rather catastrophic with $R^2$ of {{< var local_azulenes_r2_f >}} implying no correlation whatsoever ([@fig-local-azulenes]c). The situation for the $\DeltaE{ST}$ also worse than for the azaphenalenes, with $R^2=$ {{< var local_azulenes_r2_cis >}} with CIS + DSP, a marked improvement on {{< var local_azulenes_r2_scf >}} with SCF + DSP ([@fig-local-azulenes]e,f). Unfortunately, the negative gaps are not recovered, leading to recalls of {{< var local_azulenes_recall_cis >}} as none of five inverted molecules could be identified ([@tbl-local-azulenes]). It could be speculated that the worse performance for the azaazulenes comes from the fact that they are non-alternants, for which the approximations are expected to be less applicable. Considering the contrasting performance for the azaphenalenes and azaazulenes, it seems clear that the PPP screening protocol will inevitably miss some hits belonging to particular compound classes. To investigate the global performance of the model we turned to a more diverse set with 16 different compound classes.

@@

Here we used a dataset previously generated by some of us using rational design rules, comprising 16 different molecular scaffolds, the details of which will be presented elsewhere. Out of {{< var global_n_total >}} compounds, only {{< var global_n_failed >}} failed, which corresponds to a success rate of {{< var global_percent_success >}}. To put this into context, ~25% of the CASSCF calculations and 9% of the CIS calculations failed in a recent HTVS workflow by Troisi, Padula and co-workers.[@omarIdentificationUnknownInverted2023] The total runtime using four cores on the M2 processor was {{< var time_cpu_hours >}} CPU hours, with the mean time per compounds at {{< var time_mean >}} seconds. The S~1~ and T~1~ excitation energies are fairly reproduced with ([@fig-global-all]a,b) $R^2$ of {{< var global_all_r2_singlet >}} and {{< var global_all_r2_triplet >}} vs {{< var reference_level >}}, respectively. As seen in [@fig-global-all-gaps]a, the $\DeltaE{ST}$ at the SCF + DSP level shows a moderate $R^2$ of {{< var global_all_r2_scf >}}, which increases significantly to {{< var global_all_r2_cis >}} at the CIS + DSP level ([@fig-global-all-gaps]b). However, due to a systematic overestimation of $\DeltaE{ST}$, the F1 score is still low at {{< var global_all_f1_cis >}} and a poor recall of {{< var global_all_recall_cis >}} ([@tbl-global-all]). We therefore added a linear correction $\DeltaEST[DSP]{CIS,LC}=$ {{< var global_lc_coef >}} $\times\DeltaEST[DSP]{CIS}$ {{< var global_lc_intercept >}}, which increases the F1 score to {{< var global_all_f1_cis_corr >}} and the recall to {{< var global_all_recall_cis_corr >}} ([@fig-global-all-gaps]c). The specificity simultaneously decreases from {{< var global_all_specificity_cis >}} to {{< var global_all_specificity_cis_corr >}}, but overall the linear correction would be preferable for virtual screening where there is a stronger focus on finding rare hits. Applying the linear correction to the azaphenalenes ([@fig-local-corr]a) and ([@fig-local-corr]b) leads to an improvement in the recall of the former from {{< var local_azaphenalenes_recall_cis >}} to {{< var local_azaphenalenes_recall_cis_corr >}} while still none of the inverted azaazulenes are recovered. The oscillator strengths are unfortunately poorly correlated for the rational design set, with $R^2=$ {{< var global_all_r2_oscillator >}}, although the Spearman rank correlation is more encouraging at {{< var global_all_rho_oscillator >}} ([@fig-global-all]c). Taken together, the results reinforce the conclusions from the case study of the azaphenalenes, that the PPP protocol is suitable for screening $\DeltaE{ST}$ while it struggles for oscillator strengths.

A more detailed breakdown of the correlations into the different scaffolds shows that the highest $R^2$ of {{< var global_scaffolds_r2_max >}} is obtained for {{< var global_scaffolds_r2_max_name >}}, while the lowest is obtained for {{< var global_scaffolds_r2_min_name >}} with {{< var global_scaffolds_r2_min >}}. For recall, the corresponding range is {{< var global_scaffolds_recall_max >}} for {{< var global_scaffolds_recall_max_name >}} to {{< var global_scaffolds_recall_min >}} for {{< var global_scaffolds_recall_min_name >}}. The wide range in $R^2$ and recall further reinforces that the method struggles with some particular compounds classes despite its favorable global performance. Despite these problems, the method is able to capture inverted molecules in {{< var global_scaffolds_n_captured >}} out of the 15 scaffolds where they occur. Even for the azaazulenes included in this screening set at least some candidates are found, in contrast to the findings above for the azaazulenes. The per-scaffold oscillator strengths are arguably sufficiently good to allow for local screening in some cases ([@fig-global-scaffolds-oscillator]). The per-scaffold S~1~ and T~1~ excitation energies can be found in ([@fig-global-scaffolds-singlet]) and ([@fig-global-scaffolds-triplet]), respectively. Notable failures are for {{< var I >}} with $R^2$ of {{< var global_scaffolds_singlet_r2_I >}} and {{< var global_scaffolds_triplet_r2_I >}}, respectively, and {{< var XII >}} and {{< var XVI >}}, where some negative T~1~ excitation energies indicate restricted-unrestricted instabilities.

@@

# Data availability

The code to perform all the calculations and to generate the PDF of the manuscript are found at https://https://github.com/kjelljorner/ppp-invest. Snakemake[@molderSustainableDataAnalysis2021] was used as workflow manager, ensuring reproducible data generation. An archive of the workflow run used to generate this manuscript can be found on https://doi.org/[doi] together with all the output data generated.
